Ages of Planets and Stars Observed by Kepler

### 2024-04-10 paper/ms.tex
update ms: rewrite abstract

% <250 words
\begin{abstract}
  We report empirically calibrated rotation- and lithium-based ages
  for planets and stars observed by the primary Kepler mission.
  %
  Stellar rotation periods are collected from previous analyses of
  Kepler data; lithium content is assessed using both new and
  archival HIRES spectra.
  %
  We report ages for \nkoisgoldwithages\ FGK stars with transiting
  planets, with relative statistical precisions $\sigma_t / t$ of XX
  +YY -ZZ\%.  This is a factor of X.X +Y.Y -Z.Z more statistically precise than
  isochronal ages for the same stars, and $\approx$X.X times less
  precise than cluster-based ages.
  %
  The results include XXX planets younger than 1\,Gyr (at 3$\sigma$),
  and YYY planets younger than 0.5\,Gyr (at 3$\sigma$).
  %
  We also report rotation-based ages for XX,XXX stars observed
  by Kepler, aka.~the ``target star sample''.
  %
  In the latter sample of dwarf FGK stars observed by Kepler, we find
  that the gyrochrone age distribution rises linearly from 0 to
  3\,Gyr, with stars 2\,Gyr old being twice as common in the Kepler
  field as stars 1\,Gyr old.
  %
  This qualitatively agrees with previous analyses of the Milky Way's star
  formation history, though it offers greater sensitivity in the range of
  $\approx$0.1-1\,Gyr.
  %
%  The sample of detected Kepler planets has a similar age
%  distribution, though it is slightly deficient at $\lesssim$0.2\,Gyr,
%  likely due to decreased detection sensitivity.
  %
  Combining our stellar age catalog with Kepler's completeness products, we perform
  a hierarchical bayesian analysis to infer the average number of planets per star
  as a function of both stellar and planetary parameters.
  %
  While we recover many previously known correlations with stellar mass and metallicity,
  our analysis also highlights the importance of age in shaping the ``sub-Saturn desert'',
  and in influencing the evolution of the mini-Neptune and super-Earth populations.
  %TODO will this be the case?
%  We also comment on a few previously reported evolutionary trends in
%  the size evolution of close-in mini-Neptunes (2-4\,\rearth) and
%  super-Earths (1-2\,\rearth), a few of which we recover at comparable
%  ($\sim$3$\sigma$) significance to previous work, and one of which we
%  note for the first time.
\end{abstract}

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.5\textwidth]{prot_teff_Santos19_Santos21_clean0.pdf}
		}
		\subfloat{
			\includegraphics[width=0.5\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Rotation periods for Kepler stars (left) and planets (right)}.
    The periods are from \citet{Santos_2019} and \citet{Santos_2021}
    (\citetalias{Santos_2019} and \citetalias{Santos_2021}), and the
    temperatures are primarily from \citet{Berger_2020a_catalog}.  The
    gray lines are ``mean fits'' from \citet{Bouma_2023} to the slow
    rotation sequences of open clusters.  The stellar sample shows
    only apparently single stars near the main sequence with $\log
    g$$>$4.2, RUWE$<$1.4, 3800$<$$T_{\rm eff}$/K$<$6200, that are not
    flagged as eclipsing binaries.  The planet sample has the same
    stellar cuts, and shows confirmed and candidate planets from the
    cumulative KOI table, requiring them to have a MES$>$10.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

@@

%TODO: does JUMP mirror all HIRES data from KOA?
We queried a combination of the Keck Observatory Archive and the
internal CPS archive to collect all HIRES spectra that were available
for the \nplhoststarwgyroagewithgrazingandhighruwe\ stars
described in Section~\ref{subsec:tally} for which it might be possible
to derive rotation-based ages.
We imposed no constraint on whether the iodine cell was present, since
there are no relevant lines in the region of the 
\ion{Li}{1} 6708\,\AA\ doublet.
This exercise yielded at least one spectrum for \nlithiumstars\ and 
\nlithiumplanets\ planets.
These observations were acquired between 6 September 2009 and 16
May 2024, primarily by the California Kepler Survey (CKS,
CITE).
In preparation for this study, we additionally collected {\bf XX}
spectra for ``confirmed'' planets with rotation-based ages below
1\,Gyr.

>>>>

% <250 words
\begin{abstract}
  Recent studies of stars in open clusters have yielded a detailed picture
  of how the rotation rates and lithium content of FGK stars evolve
  during the main sequence.
  %
  Here we apply this knowledge to planets and stars observed by the
  primary Kepler mission (2009-2013), and report the resulting rotation and
  lithium-based ages.
  %
  Rotation periods are collected from previous analyses of
  Kepler data; lithium is measured from new and archival HIRES
  spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  and \nplwgyroagenograzing\ planets for which we believe
  gyrochronology is likely valid. %, and for many more for which the underlying assumptions of
%  the technique may be compromised.
  %
  The relative statistical precisions from rotation, $\sigma_{t_{\rm
  gyro}} / t_{\rm gyro}$, are XX +YY -ZZ\%, which is X.X +Y.Y -Z.Z more
  statistically precise than isochronal ages for the same stars, and
  $\approx$X.X times less precise than cluster-based ages.
  %
  The results include \nplyounggyrotwosigma\
  planets younger than 1\,Gyr at 2$\sigma$ (\nplyounggyrothreesigma\
  planets at 3$\sigma$).
  %
  While lithium provides only marginal added information for most of
  this sample,
  we report robust two-sided age measurements from both rotation and
  lithium for 14 confirmed $<$500\,Myr planets, including five
  mini-Neptunes, five super-Earths, three Earths and sub-Earths, and a
  Neptune-sized planet.
  %
  In the broader Kepler target star sample,
  we find
  that the gyrochrone age distribution rises linearly from 0 to
  3\,Gyr, with stars 2\,Gyr old being twice as common in the Kepler
  field as stars 1\,Gyr old.
  %
  This approach offers greater sensitivity to main sequence ages than previous analyses of the Milky Way's star
  formation history, and the resulting age distribution bounds
  the relative balance of ``young'' and ``old'' planets detectable not only by Kepler, but also
  by future wide-field transit surveys.
  %
%  The sample of detected Kepler planets has a similar age
%  distribution, though it is slightly deficient at $\lesssim$0.2\,Gyr,
%  likely due to decreased detection sensitivity.
  %
  %Combining our stellar age catalog with Kepler's completeness products, we perform
  %a hierarchical bayesian analysis to infer the average number of planets per star
  %as a function of both stellar and planetary parameters.
  %
  %While we recover many previously known correlations with stellar mass and metallicity,
  %our analysis also highlights the importance of age in shaping the ``sub-Saturn desert'',
  %and in influencing the evolution of the mini-Neptune and super-Earth populations.
  %TODO will this be the case?
%  We also comment on a few previously reported evolutionary trends in
%  the size evolution of close-in mini-Neptunes (2-4\,\rearth) and
%  super-Earths (1-2\,\rearth), a few of which we recover at comparable
%  ($\sim$3$\sigma$) significance to previous work, and one of which we
%  note for the first time.
\end{abstract}

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.5\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
      % vs prot_teff_Santos19_Santos21_clean0.pdf !
		}
		\subfloat{
			\includegraphics[width=0.5\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
    {\bf Rotation periods for Kepler stars (left) and planets (right)}.
    The periods are from \citet{Santos_2019} and \citet{Santos_2021}
    (\citetalias{Santos_2019} and \citetalias{Santos_2021}), and the
    temperatures are primarily from \citet{Berger_2020a_catalog}.  The
    gray lines are ``mean fits'' from \citet{Bouma_2023} to the slow
    rotation sequences of open clusters.  The stellar sample shows
    only apparently single stars near the main sequence with $\log
    g$$>$4.2, RUWE$<$1.4, 3800$<$$T_{\rm eff}$/K$<$6200, that are not
    flagged as eclipsing binaries.  The planet sample has the same
    stellar cuts, and shows confirmed and candidate planets from the
    cumulative KOI table, requiring them to have a MES$>$10.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

@@

%TODO: does JUMP mirror all HIRES data from KOA?
We queried the internal CPS archive to collect all HIRES spectra that
were available for the \nplhoststarwgyroagewithgrazingandhighruwe\
stars for which rotation-based ages can be derived.
This exercise yielded at least one spectrum for \nlithiumstars\ and 
\nlithiumplanets\ planets.
We did not constrain whether the iodine cell was used, since
there are no relevant lines in the region of the 
\ion{Li}{1} 6708\,\AA\ doublet.
These observations were acquired between 6 September 2009 and 16
May 2024, primarily by the California Kepler Survey (CKS,
CITE).
In preparation for this study, we additionally collected {\bf XX}
spectra for ``confirmed'' planets with rotation-based ages below
1\,Gyr.
